An Efficient Interpolation Technique for Jump Proposals in Reversible-Jump Markov Chain Monte Carlo Calculations

### 2015-04-06 paper.tex
Implemented Ewan's suggestions.

Markov chain Monte Carlo (MCMC) methods attempt to stochastically
produce parameter samples with density proportional to the posterior
probability distribution.  In MCMC techniques, the primary target is
an accurate estimate of the posterior distribution.  (We note that an
alternative stochastic method for exploring a model parameter space,
nested sampling \cite{Skilling:2004,Skilling:2006,Feroz:2009},
focuses on evidence computation rather than sampling the posterior
probability density functions.)  It is not straightforward to compute
the model evidence from MCMC samples.  The most direct way to estimate
the evidence for a model from MCMC samples is to compute the
harmonic-mean estimator, but this estimator of the evidence suffers
from infinite variance
\cite{NewtonRaftery:1994,Chib:1995,vanHaasteren:2009}.  MCMC
implementations with parallel tempering \cite{SwendsenWang:1986,EarlDeem:2005} allow
for evidence computation via thermodynamic integration, but these can
be computationally costly.

@@

When the goal is model selection between several known models, only
the \emph{relative} evidence of each model is needed.  In this
circumstance, the Reversible Jump MCMC technique first introduced in
\cite{Green1995} is one of the most reliable and accurate ways to
compare the models.  Reversible Jump MCMC (RJMCMC), described more
fully in Section \ref{sec:reversible-jump}, performs a standard MCMC
in a superspace that is a direct sum of all the model parameter
spaces.  Such an MCMC involves both intra- and inter-model jumps; the
number of MCMC samples in each model's parameter space is proportional
to that model's relative evidence in the suite of models being
compared.

@@

A Markov chain has the property that the probability distribution of
the next state can depend only on the current state, not on the past
history: 
%
\be 
p(\vtheta^{(j+1)})=\int_{V} d\vtheta^{(j)} p(\vtheta^{(j)} \to
\vtheta^{(j+1)}) p(\vtheta^{(j)}), \ee
%
where the jump probability $p(\vtheta^{(j)} \to \vtheta^{(j+1)})$
depends only on $\vtheta^{(j)}$ and $\vtheta^{(j+1)}$.  An additional
requirement for an MCMC arises from the fact that the desired
distribution is the equilibrium distribution.  In other words, if we
assume that state $(j)$ of the chain is sampled from the desired PDF,
$p(\vtheta^{(j)})=p(\vtheta^{(j)}|d,M)$ , then the next state $(j+1)$ must
be sampled from the PDF as well, so that
$p(\vtheta^{(j+1)})=p(\vtheta^{(j+1)}|d,M)$; this condition is known as
``detailed balance.''

@@

The samples produced by an MCMC algorithm can be used to directly
perform a Monte Carlo evidence integral.  This results in a harmonic
mean estimator for the evidence, which suffers from a large variance
and bias \cite{NewtonRaftery:1994,Chib:1995,vanHaasteren:2009}.
Additional techniques for the direct integration of evidence, also
based on a kD tree decomposition of the parameter space (see
Sec.~\ref{sec:kDTree}), are described in \cite{Weinberg2009}.  These
techniques are promising, but in some cases suffer 
from large variance and bias \cite{Farr2010}.  An alternative approach to model
selection among a set of models is based on performing an MCMC in a
``super-model'' that encompasses all of the models under
consideration; this is known the the Reversible Jump Markov chain
Monte Carlo (RJMCMC).

@@

We are grateful to Neil Cornish for interesting discussions.  IM
acknowledges support from the NSF Astronomy and Astrophysics
Postdoctoral Fellowship, award AST-0901985, for the early stages of this work.  
WF acknowledges support
from NSF grant AST0908930 and NASA grant NNX09AJ56G. DS acknowledges
support from a Northwestern University Summer Undergraduate Research Grant.  IM is grateful for the hospitality offered by Monash University and, particularly, Yuri Levin.  This work has been partially supported by the UK Science and Technology Facilities Council.

>>>>

Markov chain Monte Carlo (MCMC) methods attempt to stochastically
produce parameter samples with density proportional to the posterior
probability distribution.  In MCMC techniques, the primary target is
an accurate estimate of the posterior distribution.  (We note that an
alternative stochastic method for exploring a model parameter space,
nested sampling \cite{Skilling:2004,Skilling:2006,Feroz:2009},
focuses on evidence computation rather than sampling the posterior
probability density functions.)  It is not straightforward to compute
the model evidence from MCMC samples.  The most direct way to estimate
the evidence for a model from MCMC samples is to compute the
harmonic-mean estimator, but this estimator of the evidence  can suffer
from infinite variance 
\cite{NewtonRaftery:1994,Chib:1995,vanHaasteren:2009,WolpertSchmidler:2012}.  MCMC
implementations with parallel tempering \cite{SwendsenWang:1986,EarlDeem:2005} allow
for evidence computation via thermodynamic integration \cite{FrielPettitt:2008}, but these can
be computationally costly.

@@

When the goal is model selection between several known models, only
the \emph{relative} evidence of each model is needed.  In this
circumstance, the Reversible Jump MCMC technique first introduced in
\cite{Green1995} is one of the most reliable and accurate ways to
compare the models.  Reversible Jump MCMC (RJMCMC), described more
fully in Section \ref{sec:reversible-jump}, performs a standard MCMC
in space that is an augmented union of all the model parameter
spaces.  Such an MCMC involves both intra- and inter-model jumps; the
number of MCMC samples in each model's parameter space is proportional
to that model's relative evidence in the suite of models being
compared.

@@

A Markov chain has the property that the probability distribution of
the next state can depend only on the current state, not on the past
history: 
%
\be 
p(\vtheta^{(j+1)})=\int_{V} d\vtheta^{(j)} p(\vtheta^{(j)} \to
\vtheta^{(j+1)}) p(\vtheta^{(j)}), \ee
%
where the jump probability $p(\vtheta^{(j)} \to \vtheta^{(j+1)})$
depends only on $\vtheta^{(j)}$ and $\vtheta^{(j+1)}$.  An additional
requirement for an MCMC arises from the fact that the desired
distribution is the equilibrium distribution.  Detailed balance requires that
$p(\vtheta^{(i)}) p(\vtheta^{(i)} \to \vtheta^{(j)}) = p(\vtheta^{(j)}) p(\vtheta^{(j)} \to \vtheta^{(i)})$.

% In other words, if we assume that state $(j)$ of the chain is sampled from the desired PDF, $p(\vtheta^{(j)})=p(\vtheta^{(j)}|d,M)$ , then the next state $(j+1)$ must be sampled from the PDF as well, so that $p(\vtheta^{(j+1)})=p(\vtheta^{(j+1)}|d,M)$; this condition is known as ``detailed balance.''

@@

The samples produced by an MCMC algorithm can be used to directly
perform a Monte Carlo evidence integral.  This results in a harmonic
mean estimator for the evidence, which may suffer from infinite variance \cite{NewtonRaftery:1994,Chib:1995,vanHaasteren:2009,WolpertSchmidler:2012}.
Additional techniques for the direct integration of evidence, also
based on a kD tree decomposition of the parameter space (see
Sec.~\ref{sec:kDTree}), are described in \cite{Weinberg2009}.  These
techniques are promising, but in some cases suffer 
from large variance and bias \cite{Farr2010}.  An alternative approach to model
selection among a set of models is based on performing an MCMC in a
``super-model'' that encompasses all of the models under
consideration; this is known the the Reversible Jump Markov chain
Monte Carlo (RJMCMC).

@@

We are grateful to Neil Cornish for interesting discussions.  WF acknowledges support
from NSF grant AST0908930 and NASA grant NNX09AJ56G.  IM
acknowledges support from the NSF Astronomy and Astrophysics
Postdoctoral Fellowship, award AST-0901985, for the early stages of this work.  
 DS acknowledges
support from a Northwestern University Summer Undergraduate Research Grant. IM is grateful for the hospitality of the Monash Center for Astrophysics supported by a Monash Research Acceleration Grant (PI Y.~Levin).  This work has been partially supported by the UK Science and Technology Facilities Council.
